# Introduction {#intro-chapter}
```{r cache = TRUE, echo=FALSE, warning=FALSE, eval=TRUE, message=F}
# Chunk defaults
knitr::opts_chunk$set(cache = TRUE, message = FALSE, warning=FALSE, eval=TRUE, echo=FALSE, 
                      error=FALSE,
                      # fig.width=7, fig.height=7,
                      fig.align='center',
                      out.width = "85%")
```


Anthropogenic activity in the last few decades will continue to influence the interations within and among ecological systems worldwide. The complexity of and drivers of changes in coupled human-natural systems is consequently altered, further limiting our ability to detect and predict change and impacts of change [@liu_complexity_2007; @scheffer_critical_2009]. Early warning systems are developed to detect,  and in some cases predict, abrupt changes in disparate systems [e.g. cyber security [@], infrastructure [@], banking crises [@davis_comparing_2008], and agricultural systems]. The need to develop and improve early warning systems for natural and coupled human-natural systems is exacerbated by the consequences of climate change and globalization, especially when the human-related stakes are high. 

Forecasting undesirable change is, arguably, the holy grail of ecology. Paired with an understanding of system interactions, a forecast isideal if it provides information with sufficient time to prevent or mitigate unwanted systemic change. Early warning systems (or early warning signals, or early warning indicators) have been developed and tested for some ecological systems data (especially marine fisheries time series and for nutrient loading in shallow lakes). Despite the quantitative methods proposed as early warning systems for ecological data (hereafter referred to as regime detection measures, RDMs), most are currently of limited practical utility. This paradox may be a consequence of existing ecological early warning systems (or other quantitative methods for identifying systemic change)  having one or more of the following characteristics:

  1. not generalizable across systems or system types (especially when it requires a model or a determinsitic function to describe the system) 
  1. require a large number of observations  
  1. difficult to implement   
  1. difficult or to interpret   
  1. requires an understanding of the drivers of change  
  1. performs poorly under uncertainty
  1. give no uncertaintiy around estimates (tying into interpretation issues)  
  1. cannot handle noisy data
  1. ignores or does not sufficiently account for observation error  
  1. no baseline with which to compare results
  1. no application/testing on empirical systems data

Research focusing on these areas as they relate to RDMs will contribute to the advancement and improvement of existing early warning systems, and will, hopefully, highlight methods which are useful and which are not to practitioners and decision makers. 

The overarching aim of this dissertation is to advance our understanding of the utility and limitations of select early warning systems. Specifically, I focus on RDMs capable of analyzing multi-varaible data, including temporally- and spatially-explicit. Although the most widely-applied RDMs proposed in the ecological literature are those deveoped for and tested on single-variable time series (e.g., temperature or fisheries stock time series), the utility of these methods in multi-variable systems (data) is limited. Regime detection metrics for tracking and identifying changes in multivariable systems data are of greater use than single-variable RDMs in systems within which a change manifests dynamically and across multiple variables (e.g., species). Multivariable RDMs may also prove advantageous when the drivers of systemic change are unknown. Further, ecological systems are noisy, and ecological systems data are messy. 

### Dissertation structure

The chapters hereain are written as separate, publishable manuscripts. Where applicable, co-authors are listed in the front matter of the chapter. The dissertation comprises an introduction to the dissertation (Chapter \@ref(intro-chapter)}), a brief overview of early warning systems (or regime detection measures) for ecological systems data (Chapter \@ref(rdmReview-chapter)), a detailed guide to Fisher Information as a RDM written for ecologists (Chapter \@ref(fiGuide-chapter)), an application of Fisher Information to spatially-explicit data (Chapter \@ref(fisherSpatial-chapter)),  introduction of 'new' RDM, Distance Travelled (Chapter \@ref(distTravelled-chapter)), a study of data quality and data loss on select RDMs including Distance Travelled and Fisher Information (Chapter \@ref(sims-chapter)), and conclusions (Chapter \@ref(conclusions-chapter)). 

## Glossary
Research surrounding regime shifts, threshold identification, change-point detection, bifurcation theory, etc. is muddled with jargon. Here, I provide a glossary (Table \@ref(glossary)) for terms and concepts that may either be unfamiliar to the practical ecologist, or may have multiple meanings among and within ecological researchers and practitioners. 

```{r glossary, echo = F, eval = F, message=F, warning=F}
# Create a table of definitions 
require(kableExtra)

glossary <- read.csv(paste0(here::here(), 
                            "/chapterFiles/glossary.csv"))

kable(glossary, "latex", caption = "A table of definitions for terms, theories, and phrases often appearing in ecological regime shift literature.") %>% 
kable_styling(full_width = F) %>%
kableExtra::column_spec(1, bold = T)
```




