# Robustness of Multivariate Regime Detection Measures to Varying Data Quality and Quantity {#resampling}

```{r cache = TRUE, echo=FALSE, warning=FALSE, eval=TRUE, message=FALSE, results="hide"}
# Chunk defaults
# require(knitr)
knitr::opts_chunk$set(cache = TRUE, cache.path = "_cache/", message = FALSE, warning=FALSE, eval=TRUE, echo=FALSE, error=FALSE, fig.align='center', out.width = "85%")
figDir <- here::here("chapterFiles/resampling/figsCalledInDiss/")
loadFig<-function(fn){paste0(figDir,fn,".png")}
```
## Introduction 
Ecological systems have many unpredictable and variably interacting components [@jorgensen_new_2011]. Methods for analyzing these complex systems, e.g. Dynamic Bayesian Networks, network models, and food webs are designed to handle these complexities, yet require data- and knowledge-intensive models. Although ecological data collection and data management techniques are improving [@lasorte2018opportunities], the aforementioned approaches to modeling and understanding complex system are often infeasible in ecosystem research and management [@clements_including_2016]. 

A growing concern with anthropogenic impacts on the environment has increased the demand for mathematical and statistical techniques that capture these dynamics. These often undesirable changes in the structure or functioning of ecological systems are often referred to as *regime shifts*, *regime changes*, *state change*, *abrupt change*, etc. [@andersen_ecological_2009] . A yet-unattained goal of ecological research and management is to reach a point where these methods can predict impending regime shifts in real-time and with high confidence. Ideally, ecological regime shift detection methods (hereafter, regime detection measures) would require little knowledge of the intrinsic drivers of the system, and the users of the method would not be required to know if and where a regime shift occurred in the data. 

Despite the suite of regime detection measures in the environmental and ecological research literatures, they are not used in ecological management. We can describe the current state of regime detection measures as being either system specific (i.e., the method is not widely applicable or generalizable across systems) or not. Methods of the latter type are convenient in that they can be applied across various system and data types, but the results of these analyses require some degree of subjective interpretation [@clements2018indicators; *c.f.* @batt2013changes]. Efforts to develop and/or improve regime detection measures that can handle these biases will aid the advance of regime detection measures research and application. 

Current efforts to improve regime detection measures may be stunted by the lack of application beyond simple and/or theoretical (toy) systems data. Like most statistical and mathematical approaches, the evolution of many regime detection measures begins with application to theoretical data, followed by application to empirical data. Current applications of regime detection measures to empirical, ecological data are largely limited to data describing populations [@anderson_community_1999;@alheit_synchronous_2005; @deyoung_regime_2008], climatic, marine, and Paleolithic regime shifts [@spanbauer_prolonged_2014; @kong2017hydrological; @yang_10_2006], with few applications to terrestrial data [*c.f.* @bahlai2015shifts; @sundstrom2017detecting]. Although testing the performance and inference boundaries of theoretical and simple systems is important, they are of little use to ecosystem managers if they are not proven to be easily and reliably applicable to their system. Additionally, regime detection measures should be capable of handling empirical ecological data are often sparse and noisy.

Ecological systems data is not only expensive to capture, but are often difficult to perfectly capture due to the large  process and observation errors. The variability resulting from imperfect observation influences data quality and quantity, sometimes limiting the potential numerical tools used to identify trends and changes in the system in question [@thrush2009forecasting]. Some methods, new and old, are proposed in the literature as regime detection measures which are capable of handling data limitation and quality issues inherent in ecological data and require few subjective decisions for choosing state variables and interpreting results. For example, variable reduction techniques, e.g. principal components analysis [@rodionov_application_2005; @andersen_ecological_2009; @reid_global_2016] and clustering algorithms [@weijerman2005regime;@weissman2016predicting], an index of variance [@brock_variance_2006]  and Fisher Information [@cabezas_towards_2002; @fath_exergy_2004; @karunanithi_detection_2008] were introduced as methods which collapse the system into a single indicator of ecological regime shifts. Although these methods have been tested on empirical ecological systems data, their robustness to empirical data quality and quantity have yet to be examined. 
In this Chapter I examine the influence of observation and process errors on the inference obtained from select multivariable regime detection measures. There are two major objectives:  
1. Identify the effects of data quality on regime detection measure inference.  
1. Identify the effects of data quantity on regime detection measure inference. 
1. Explore the relative performance of velocity (described in Chapter \@ref(velocity)) to the abovementioned methods under multiple scenarios.   

This Chapter provides baseline relative performance estimates of select, multivariable regime detection measures under various scenarios of data quality and quantity. The results from this Chapter inform the practical ecologist of the potential limitations to consider when applying these regime detection measures to their data, and has potential to inform the data collection process. Additionally, the software accompanying this Chapter allows the end user to implement these methods on this diatom system, a toy system, or their own data.

## Data and Methodology
### Study system and data  
I used paleodiatom time series from a freshwater system in North America (Foy Lake, present day Montana) that apparently underwent a rapid shift in algal community dynamics at multiple periods in time. This datum comprises a single soil core sample, from which the relative abundances of 109 diatom species were identified at  768 observations (time points) over $\approx7,000$ years (Figure \@ref(fig:origDat). Althouh the soil core was sampled at regular distances, the soil accumulation process is not necessarily linear over time, resulting in irregularly-sampled observations (i.e., time elapsed between sampling points differs varies; see Figure \@ref(fig:timeElapsed)). This datum was published in @spanbauer_prolonged_2014 and can be downloaded at the publisher's website.

```{r origDat, fig.align='center', out.width = "85%",  echo=FALSE,  fig.cap="Relative abundances of the diatom species in Foy Lake over the time period."}
knitr::include_graphics(loadFig("origDataRelAbundance"))
```

```{r timeElapsed, fig.align='center', out.width = "85%", echo=FALSE, fig.cap="The amount of time elapsed between observations."}
knitr::include_graphics(loadFig("timeElapsed"))
```

### Regime detection measures
Fewer model-free regime detection metrics exist than do model-based metrics (Chapter \@ref(rdmReview)) and of these, only a few are suggested for handling multivariable data. Here, I examine the regime detection metrics that are model-free and can handle multivariable data: velocity (Chapter \@ref(velocity)), the Variance Index [@brock_variance_2006] and Fisher Information. These methods and the primary sources are described below.

#### Velocity ($v$) 
In Chapter \@ref(velocity), I describe a new method, __velocity__, $v$, as a potential dimension reduction and regime detection method. First introduced in by @fath_regime_2003 as one of multiple steps in calculating their variant of Fisher Information, velocity calculates the cumulative sum of the square root of the sum of the squared change in all state variables over a period of time (Eq. \@ref(eq:velocityEq)). Steps for calculating this metric are described in detail in Chapters \@ref(fiGuide) and \@ref(velocity).  

\begin{equation}
\begin{array}{rcr}
\Delta s_i = \sqrt{\sum_{j=1}^{n} (x_{i,j} -x_{i-1, j})^2}
s_k =  \sum_{i=2}^{k}\Delta{s_i}
2\leq k \leq n
v =\frac{\Delta s}{\Delta t}  
\end{array}
(\#eq:velocityEq)
\end{equation}

#### Variance Index  
The Variance Index was introduced by @brock_variance_2006, and is simply defined asthe maximum eigenvalue of the covariance matrix of the system over some period (window) of time. The Variance Index (also called Variance Indicator) was originally applied to a modelled system [@brock_variance_2006], and has since been applied to empirical data [@spanbauer_prolonged_2014; @sundstrom2017detecting]. Although rising variance has been shown to manifest prior to abrupt shifts in some empirical systems data  [@van2005implications;@brock_variance_2006], the Variance Index, which is intended for multivariate data,  appears most useful when the system exhibits a discontinuous (non-linear)  shift [@brock_variance_2006].

#### Fisher Information    
Fisher Information ($I$) is essentially calculted as the area under the curve of the acceleration to the fourth degree ($s''^4$) divided by the squared velocity ($s'^2$; also referred to as $v$ in Chapter \@ref(velocity)) of the distance travelled by the system, $s$ over some period of time ($T$), and is given in Eq. \@ref(eq:fiDerivs2):  

\begin{equation}   
    I = \frac{1}{T} \int_0^T dt\left[\frac{s''^2}{s'^4}\right]^2 \\  
  (\#eq:fiDerivs2)  
\end{equation} 

I describe this method in complete detail in Chapter \@ref(fiGuide).

#### Using moving window analysis to calculate  Fisher Information and Variance Index
Unlike $velocity$, the Variance Index and Fisher Information are calculated using moving window analysis. That is, over the entire time series, $T^*$, these metrics are calculated within multiple windows of time, $T$. In this approach, all state variables, $x_i$, are used to inform the calculations (of Variance Index and Fisher Information) over a time interval, $T$, where $T$ is the length in [time] units of the time interval and satisfies the following conditions: $T < T^*$ and  $2\leq T < (T^*-1)$. If $T = T^*-1$, then only a single value of the metrics will be calculated for entire time series, which does not allow for any estimate of change. 

When using these metrics in the context of identifying abrupt changes in ecological systems data across $T*$, it is ideal the value of $T$ meets the following conditions: $3 < T \ll T^*-1$. The length of a time window dictates the number of calculations one can obtain over $T^*$, such that the number of potential metric calulations increases as $\frac{T}{\ T^*}$ decreases. Previous applications of moving window analyses to calculate Fisher Information found that at least eight observations (time points) should be used. 

An additional parameter is required when conducting moving window analyses: the amount of time points by which the window advances. In order to maximize the data, I force the window to advance at a rate of one time unit. However, it is important to note that because these data are not sampled annually and the because the window always advances by a single time unit, the number of observations included in each calculation will not be the same. If fewer than 5 observations are in a window, I did not calculate metrics, advancing the window forward. 
I assigned the calcuated values of Fisher Information and Variance Index within each moving window to the __end__ (the last time unit) of the moving window. I  temporal analyses, assigning the value to any other point in time (e.g., the beginning or the middle) muddles the interpretation of the metric over $T^*$. Also note that this method has the potential to result in calculating a metric for all integers between $0.20 T^*$ and $T^*$.

### Resampling Techniques for Simulating Data Quality and Quantity Issues
Using a bootstrap approach I calculated the regime detection measures over varying degrees of scenarios to simulate data quality and data quantity issues that are common to ecological data anlaysis. The scnearios are categorized as *observations* and *species*. The observations scenario simulates a loss of temporal observations (decreasing the number of times the system was observed), and the species scenario simulates a loss of information about the system by removing a larger proportion of the species. The loss of temporal observations and the loss of species were examined at three proportions: $\textbf{P} = [0.25, 0.50, 0.75, 1.00]$, where $\textbf{P}$ is the proportion of species and time points **retained** for analysis. For example, when $\textbf{P} = 0.25$, a random selection of $25\%$ of the species are retained for analysis in the species scenario. I bootstrapped the datum over 10,000 iterations for each scenario and $\textbf{P}$ combination. Note that because when $\textbf{P} = 1.00$, all data are retained. Therefore, no resampling was conducted at this level because only a single metric (e.g. Velocity) value is possible.

Interpretation of the regime detection measures used in this analysis are currently limited to visual inspection. Therefore, I limit inference in this study largely to the impact of data loss on the variability with a regime detection measure (i.e. how robust is the measure to data loss).

## Results
In many cases the standard deviation of FI far exceeded the mean value of FI. I calculated the coefficient of variation, $\frac{\sigma}{\mu}$, for each ????!!!!??? as the proportion of data (and observations) used decreased. For example, when when we retained 

### Velocity of the distance travelled produces similar results with information 

### Fisher Information is highly sensitive to information loss 
When we bootstrap 25% of the species, the ratio of mean Fisher Information to standard deviation of Fisher Information (over $10,000$ iterations) is always $< 1$, suggesting Fisher Information does not produce fidel results when information is lost about the system. 

```{r results, out.width = "85%",  echo=FALSE,  fig.cap="Density plot of the coefficient of variation (CV) as a percentage (%) of the Fisher  Information bootstrapped samples (10,000 iterations). Densities based on all values of CV, but values >100% are not printed."}
knitr::include_graphics(loadFig("FI_cvDensity"))
```

## Discussion 
The primary results from this study confirm the sensitivity of regime detection metrics to data quantity and quality. Previous studies of the robustness of various methods have found that varinace (standard deviation, coefficient of variation) are unreliable measures of 'regime shifts' under numerous conditions.

```{r loessEx, out.width = "85%",  echo=FALSE,  fig.cap="Local regression (loess) smoothing of a dominant species in the paleodiatom community, \textit{Anomoeoneis costata} varies wildly with the span paramter. Automatic loess regressions ex "}
knitr::include_graphics(loadFig("FI_cvDensity"))
```

Some studies detrend the original time series prior to data aggregation and calculation of regime detection metrics. I did not detrend the original data for two reasons. First, the authors of the original paper analysing this dataset [@spanbauer_prolonged_2014] did not detrend species time series. Like @spanbauer_prolonged_2014 I only scaled the original data, rather than detrending. Second, detrending a time series requires yet another subjective decision by the data analyst. For example, a "spanning" parameter must be chosen when detrending (smoothing) non-linear time series using local regression (Loess) regression [@]. Finally, this data exhibits rapid and drastic shifts in community composition *and* contains a disproprtionate amount of dominant versus non-dominant species. Consequently, most species contain more zero than non-zero observations, which makes loess smooothing difficult.


Although this chapter concerns impacts of data quality and quantity based on hypothetical data collection and analytical decisions, adding yet another parameter necessitates another layer of compartive analysis. Future work studying the impact of detrending, data scaling, outlier removal, and related decisions would be of value to understanding the efficacy of these and other regime detection measures in real-world situations.  

This is not the first study to identify the sensitivity of early warning indicators and ecological regime detection metrics. 


## Ackowledgements 
This study was conceptualized at the International Institute for Applied Systems Analysis (IIASA) as part of the Young Scholars Summer Program in 2018. I thank my IIASA program supervisors, Drs. Brian Fath and Elena Rovenskaya, for their advisement of this project and IIASA scientists Drs. Matthias Jonas, Chai Molina, Piotr Zebrowski for additional support. 