---
bibliography: FI_bib.bib
output: 
  pdf_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction 

This appendix provides documentation and R code association with the paleodiatom community example. 

## Setup
### Load required packages 
You will need to install the following packages if they are not already.

```{r, echo = F}
library(tidyverse)
library(ggplot2)
library(ggthemes)
library(RColorBrewer)
library(ggjoy)
library(kedd)
library(caTools)
library(gridExtra)

# Set default ggplot theme
theme_set(theme_minimal())
```

### Load the data

Pull in the data from the supplementary materials for Spanbauer *et al.* (2014) [@spanbauer_prolonged_2014]. This data contains the percent abundances of diatom species from Foy Lake. Spanbauer *et al.* (2014) calculated the number of relative diatom valves in each sample. They removed time steps with no diatom data, claiming poor preservation rather than zero abundance. The authors also averages time steps 301 - 312.

```{r, echo = T}
data <- read_csv("https://doi.org/10.1371/journal.pone.0108936.s001")
```

## Calculate the distance metric

### Calculate the distance 
Calculate the difference, $\Delta x$, in each species' relative abundance from time *n* to time *n + 1*. Calculate the change in distance, $\Delta s$, as the sum of the squares of the change in each species. Calculate the distance as the cummulative sums of the change in distance.
```{r}
distance <- 
  data %>% 
  select(-Sample) %>% 
  arrange(YB1950) %>% 
  mutate_at(funs(lead(.)-.), .vars = vars(-YB1950)) %>% 
  gather(species, dx, -YB1950) %>% 
  group_by(YB1950) %>% 
  summarize(ds = sqrt(sum(dx^2))) %>% 
  mutate(s = cumsum(ds)) %>% 
  filter(!is.na(s))
```

Load the numerical differentiation results data. These finite differences were approximated in MatLab, using code that implements the methods found in Chartrand, R. Numerical differentiation of noisy, nonsmooth data. (2011) ISRN Applied Mathematics, Vol. 2011. Article ID 164564 [@chartrand_numerical_2011].
```{r}
# Load csv file
numDiff <- read.csv("numerical_diff_results.csv", header = F) 
names(numDiff) <- c("t","dsdt","d2sdt2")

# Create vectors of for plotting purposes
tt <- numDiff$t
vv <- numDiff$dsdt
aa <- numDiff$d2sdt
```

### Dataframe processing
Create a dataset in 'long' form for future plotting
```{r, echo = T}
longData <- 
  data %>% 
  select(-Sample) %>% 
  gather(species, relAbund, -YB1950) 

# Find some species that have "high" abundances 
topSpecies <- 
  longData %>% 
  group_by(species) %>% 
  summarize(q95 = quantile(relAbund, 0.95)) %>% 
  arrange(desc(q95)) %>% 
  filter(row_number() <= 10)
```

## Plotting
### Plot the relative abundance data over time
```{r, echo = FALSE}

# For plotting, divide species into 10 species with "high" abundance and others
plotData <- 
  longData %>%
  mutate(species = if_else(species %in% topSpecies$species, species, "Other (n = 99)")) %>%
  mutate(species = factor(species, levels = c(sort(topSpecies$species), "Other (n = 99)"))) %>%
  group_by(YB1950, species) %>%
  summarize(relAbund = sum(relAbund))

# Generate plot
(gAbund <- 
   ggplot(data = plotData,
          aes(x = YB1950, y = relAbund, fill = species)) +
   geom_area() +
   geom_rug(sides = "b") +
   scale_fill_brewer(palette = "Paired") +
   theme(legend.position = "bottom") +
   labs(x = "year", y = "relative abundance", fill = "species"))
```

### Plot the distance travelled
```{r, echo = FALSE}
(gS <- ggplot(data = distance, aes(x = YB1950, y = s)) + 
  geom_line() +
  labs(x = "year", y = "distance") +
  xlim(min(numDiff$t), max(numDiff$t))
)
```

### Plot the velocity
```{r, echo = FALSE}
(gV <- ggplot(data = numDiff, aes(x = t, y = dsdt)) + 
  geom_line() +
  labs(x = "year", y = "velocity")) +
  xlim(min(numDiff$t), max(numDiff$t))
```


### Plot the acceleration 
```{r, echo = FALSE}
(gAcc <- ggplot(data = numDiff, aes(x = t, y = d2sdt2)) + 
  geom_line() +
  labs(x = "year", y = "acceleration") +
  xlim(min(numDiff$t), max(numDiff$t))
)
```


### Plot a histogram of distance traveled
Compare histogram of distance travelled to pdf calculated from velocity.

```{r, echo = FALSE}
# Linear interpolate distance onto equal grid used for differentiation
s <- approx(distance$YB1950, distance$s, numDiff$t, "linear")$y

# Data frame for plotting
plotData <- 
  numDiff %>% 
  cbind(s) %>% 
  mutate(TT = max(t) - min(t),
         p = (1/TT)*(1/dsdt))

(
  ggplot(data = plotData) + 
    geom_histogram(aes(x = s, y = ..density..), fill = "blue", alpha = 0.5, bins = 100) +
    geom_line(aes(x = s, y = p), size = 1.1) +
    labs(x = "distance, s", y = "probability density, p(s)")
)
```

## Moving window analysis
### Specifcy parameters for the moving window
Distance over which to move the window (in units time)
```{r, echo = TRUE}
# Distance over which to move the window (in units time)
winspace <- 50

# Size of the window (in units time)
winsize <- 500

# Start and stop points for windows
t <- distance$YB1950
winStart <- seq(min(t), max(t), by = winspace)
winStop <- winStart + winsize

# Number of windows
nWin <- length(winStart)
```

### Loop over data calculating a FI value for each window
```{r}
FI_7.12 <- numeric(length(nWin))
for(i in 1:nWin){
  df <- 
    numDiff %>%
    filter(t > winStart[i],
           t <= winStop[i]) %>%
    mutate(TT = max(t) - min(t),
         p = (1/TT)*(1/dsdt))
  
  FI_7.12[i] <- (1/df$TT[1])*trapz(df$t, df$d2sdt2^2 / df$dsdt^4)
}

```

## Plots
###Plot Fisher Information for each window

```{r FIplot}

# Plot of FI
(gFI <- ggplot(data = data_frame(winStop, FI_7.12),
       aes(x = winStop, y = FI_7.12)) +
  geom_line() +
  labs(x = "time (end of window)", 
       y = "Fisher Information") +
  xlim(min(numDiff$t), max(numDiff$t)))

# Stacked plot 
gB <- ggplotGrob(gS)
gC <- ggplotGrob(gV)
gD <- ggplotGrob(gFI)
maxWidth = grid::unit.pmax(gB$widths[2:5], gC$widths[2:5], gAcc$widths[2:5], gD$widths[2:5])
gB$widths[2:5] <- as.list(maxWidth)
gC$widths[2:5] <- as.list(maxWidth)
gD$widths[2:5] <- as.list(maxWidth)

grid.arrange(gB, gC, gD, ncol=1)
```

### Export figures
```{r}
# ggsave(
#     "figures/stackplot.png",
#     # plot = grid.arrange(gA, gB, gC, gD, ncol=1) # with relative abundance plot
#     plot = grid.arrange(gB, gC, gD, ncol = 1),
#     units = "cm",
#     dpi = 300
# )
# ggsave(
#     "figures/joyPlotFisher.png",
#     plot = joy,
#     units = "cm",
#     dpi = 300
# )
```
