---
title: "UNUSED SLIDES"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: ["css/myFonts.css", "css/myTheme.css"]
    chakra: libs/remark-latest.min.js # to load the last saved verion for offline use
    nature:
      beforeInit: "http://www.jvcasillas.com/ru_xaringan/js/ru_xaringan.js"
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      ratio: "16:7"
---


--

#### - **difficult to model mathematically**     

???
Anyone studying dynamical systems or studying ecological systems probably understands that building mathewmatical models to accurately and usefully describe empirical systems is difficult. 

Especially as we move beyonbd just a handful of state  variables.
---

# The Holy Grail of Ecology <!-- ...is prediction -->
<br>  
<br>
# **forecasting change in time to prevent or mitigate undesirable consequences**
<br>


???

- **Despite this complexity** of ecological systems, we as ecologists strive to **make predictions

- One **increasingly active line of research** in ecology is "regime shifts"

<!-- -- -->

<!-- # regime shifts -->


---

# Ecological Regime Shifts 

## **what?**  a persistent change in the structure or functioning of a system 
</br>  

???
- A **persistent change in the structure or functioning of a system**
--

## **how?** loss of negative feedback(s) maintaining the system
<br>
???
- Regime shifts occur, via a number of pathways, when a system **loses the negative feedback(s) maintaining** the system in its current state

- Again, *goal* is to predict in time to prevent something bad from happening. Pipe dream? Probably...
--

## **goal?** predict in time to prevent  
---

# Regime Shifts Detection Methods
  ###   Rising Variance/Critical Slowing Down
  ###   Variance Index (max eigenvalue of covariance matrix)
  ###   Principal Coordinates Analysis
  
--
  
  ###   **Fisher Information** 
  
--

  ### **Velocity**, $v$

???

- A few of the methods capable of handling multivariable data variance, PCA and an obscure method called, Fisher Information. 

- And its from FI where we derived a new method that I am described today, called **Velocity**

---

# Toy System
```{r, echo=FALSE, out.width="55%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/exData.png"))
```

$$ x ~ normal(\mu = 25, \sigma = 5) \textrm{ for t } > 50$$
$$x ~ normal(\mu =100, \sigma = 5) \textrm{ for t } \geq 50$$
???
So the calculations for **velocity** are quite simple, and I am goign to step through the few steps first using this simple 2-variable system, where I induced a change in mean of state variables at time t = 50. 

---

# Velocity, $\frac{\Delta s}{\Delta t}$ 
#### the linear speed of a system's trajectory (e.g., in phase space)

.pull-left[
```{r, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/exData.png"))
```
]
.pull-right[
```{r, echo=FALSE, out.width="150%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/phaseplot.png"))
```
]

???

So, our method, velocity, is basically the linear speed of a system's trajectory as it moves through phase space. 

---

# Step 1: Calculate ** $\Delta s$ **, 'distance travelled'

## $$s = \textrm{system state}$$

--
</br> 


## $$\Delta s_i = \sqrt{\sum_{j=1}^{n} (x_{i,j} -x_{i-1, j})^2}$$ 

```{r,eval=FALSE, echo=FALSE, out.width="35%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/exData_t1t2.png"))
```


---

# Step 2: Calculate ** $s$ **, **cumulative** distance travelled

### $$\Delta s_i = \sqrt{\sum_{j=1}^{n} (x_{i,j} -x_{i-1, j})^2}$$

## $$s_k =  \sum_{i=2}^{k}\Delta{s_i}$$
### $$2\leq k \leq n$$

---

# Step 3: Calculate ** $v$ ** , the linear speed of ** $s$ **
#### $$v =  \frac{\Delta s}{\Delta t}$$  

```{r, echo=FALSE, out.width="55%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/dsdtplot.png"))
```

---

```{r, echo=FALSE, out.width="60%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/velocSysEx1-1.png"))
```
</br>
```{r, echo=FALSE, out.width="60%", fig.align="center"}
knitr::include_graphics(paste0(here::here(),"/figures/velocSysEx2-1.png"))
```

---
# Potential Limitations of $v$
### $v$ increases with increasing effect size

```{r, eval=TRUE,echo=FALSE, out.width="80%"}
  knitr::include_graphics(paste0(here::here(),"/figures/simVplot1.png"))
```

???
- Going back to our toy system, the velocity increases as our total change in X_i increases 

---
# Potential Limitations of $v$
### $v$ increases with increasing effect size

```{r, eval=TRUE,echo=FALSE, out.width="80%"}
knitr::include_graphics(paste0(here::here(),"/figures/simVplot2.png"))
```

???
- But does not necessarily improve as an indicator

---


# Empirical System: Paleodiatom <!--1-->

```{r, eval=TRUE,echo=FALSE, out.width="80%"}
knitr::include_graphics(paste0(here::here(),"/figures/paleoData.png"))
```
 
Data published in Spanbauer *et al.* 2014. Prolonged Instability Prior to a Regime Shift. PLoS One
  
???

We also tested the method on paleodiatom data dating back about 8,000 years before present.
Notice here there is a high species turnover
---
# Empirical System: Paleodiatom <!--2-->

```{r, eval=TRUE,echo=FALSE, out.width="100%"}
knitr::include_graphics(paste0(here::here(),"/figures/fipaleo.png"))
```


???

We also tested the method on paleodiatom data dating back about 8,000 years before present.
Notice here there is a high species turnover
---


# Empirical System: Paleodiatom <!--3 Results-->
<!-- ## Distance travelled ** $s$ ** & Velocity ** $v$ ** identify these turnover events -->

```{r, eval=TRUE,echo=FALSE, out.width="100%"}
knitr::include_graphics(paste0(here::here(),"/figures/paleoVelocity.png"))
```

???
S and V clearly identify these turnover events, but V is somewhat noisy.

---

# Smoothing Noisy Data Before Calculating ** $v$ ** 

#### Total Variation Regularized Numerical Differentiation

```{r, eval=TRUE,echo=FALSE, out.width="90%"}
knitr::include_graphics(paste0(here::here(),"/figures/paleoV.png"))
```

???
Here we have used regularized differentiation to smooth out the original data, and which in turn smooths out the velocity. 

---

# $v$ Identifies Potential Periodicities in Certain Regimes
```{r, eval=TRUE,echo=FALSE, out.width="90%"}
knitr::include_graphics(paste0(here::here(),"/figures/paleoV2.png"))
```

???
- Additioanlly, we found it interesting that velocity also identifies interesting dynamics within the system regimes. Note the periodicities in regime 1 and 2. 
- These periodicities were not discussed in origianl examinations of this data.
---

# Impacts of Changes in Data Quality on $s$ and $v$  <!-- 1 -->
```{r, eval=TRUE,echo=FALSE, out.width="90%"}
knitr::include_graphics(paste0(here::here(),"/figures/distance_95PI_s_observations.png"))
```

---
# Impacts of Changes in Data Quality on $s$ and $v$  <!-- 2 -->
```{r, eval=TRUE,echo=FALSE, out.width="90%"}
knitr::include_graphics(paste0(here::here(),"/figures/distance_dsdt_observations.png"))
```

---

# Conclusions 

### Reduction of high dimensional data
### Capable of handling noisy data
### Is not sensitive to data quality issues common in ecology
### Best when mean > variance
### Identifies regime shifts known *a priori* 

---


# Next Steps
### Numerical identification of change point in $v$  
### Compare to distance-based metrics (feedback?)
### Compare to ordination techniques
### Predictive capacity

???
- One of the primary motivations for this project was to identify an objective, intuitive, and easy-to-calculate method for identifying ecological regime shifts


---
# Related Software
### R packages (dev versions):
####  `trashbirdecology/distanceTravelled` (calculate $s$ and $v$)
####  `trashbirdecology/regimeDetectionMeasures` (calculate multiple regime detection metrics)
####  `trashbirdecology/bbsRDM` (application to spatial data)
####  `natbprice/tvdiff` (regularized differentiation)

```{r, eval=FALSE, echo=FALSE}
source(paste0(here::here(), "/R/savePrezToPDF.R"))
```